---
title: "DSA302 Project Final"
author: "G2 Group 4"
date: "11/8/2020"
output: word_document
---
Part 1
```{r}
#read the data
ZCBP=read.csv("C:/Users/Austin Lim/Desktop/SMU/AY2021/DSA302/Project/ZCBP.txt", sep="")

#exploratory data analysis
table(is.na(ZCBP))
summary(ZCBP)
head(ZCBP,5)

#prepare the data
ZCBP=dplyr::arrange(ZCBP,time)
```
(1)Plotting of zero coupon bond prices against the maturities of the bonds
```{r}
plot(ZCBP$time,ZCBP$price,
     main = "Plot of Zero Coupon Bond (ZCB) Prices against their Maturity",
     xlab = "Maturity of ZCB",
     ylab = "Price of ZCB")
```
(2)Use equation 3 to generate the forward rates
```{r}
n=nrow(ZCBP)
f0=-log(0.01*ZCBP$price[1])/ZCBP$time[1]
frate=-1*(log(ZCBP[2:n,-1])-log(ZCBP[1:(n-1),-1]))/(ZCBP[2:n,1]-ZCBP[1:(n-1),1])
frate=c(f0,frate)

plot(ZCBP$time,frate,
     main = "Plot of Forward Rates against Maturity",
     xlab = "Maturity of ZCB",
     ylab = "Forward Rates")
```
(3)Smoothing the forward rates using optim()
```{r}
# write a function to calculate the sum of squares
sumSqMin1 <- function(par, data) {
  sum((frate-(par[1]+par[2]*data[,1]+par[3]*(data[,1])^2))^2)
}
sumSqMin2 <- function(par, data) {
  sum((frate-(par[1]+par[2]*data[,1]+par[3]*(data[,1])^2+par[4]*(data[,1])^3))^2)
}
start1=c(0,0,0)
start2=c(0,0,0,0)
opt1 = optim(start1,sumSqMin1,data=ZCBP,hessian=T,method="Nelder-Mead")
opt2 = optim(start2,sumSqMin2,data=ZCBP,hessian=T,method="Nelder-Mead")

frquad = opt1$par[1]+opt1$par[2]*ZCBP$time+opt1$par[3]*(ZCBP$time)^2
frcubic = opt2$par[1]+opt2$par[2]*ZCBP$time+opt2$par[3]*(ZCBP$time)^2+opt2$par[4]*(ZCBP$time)^3

plot(ZCBP$time,frate,
     main = "Plot of Forward Rates against Maturity",
     xlab = "Maturity of ZCB",
     ylab = "Forward Rates")
lines(ZCBP$time,frquad, col="red",lwd=2)
lines(ZCBP$time,frcubic, col="green",lwd=2)
legend("bottomleft",c("empirical","quadratic","cubic"),lty=c(1,1,1),col=c("black","red","green"))
```
(4)Estimate the empirical spot rates for t in (t1, tn) using equation (4)
```{r}
breaks=c(0,as.numeric(ZCBP$time))
grid=seq(0.37,29.12,0.01)
srate=NULL

for(i in 1:length(grid)){
  srate[i]=(max(cumsum(frate[1:(findInterval(grid[i],breaks)-1)]*diff(breaks[1:findInterval(grid[i],breaks)])))+frate[findInterval(grid[i],breaks)]*(grid[i]-breaks[findInterval(grid[i],breaks)]))/grid[i]
}

grid=as.data.frame(grid)
plot(grid$grid,srate,type = "l",
     main = "Plot of Spot Rates against Maturity",
     xlab = "Maturity of ZCB",
     ylab = "Spot Rates")
```
(5)Smoothing the spot rates using optim()
```{r}
sumSqMin3 <- function(par, data) {
  sum((srate-(par[1]+par[2]*data[,1]+par[3]*(data[,1])^2))^2)
}
sumSqMin4 <- function(par, data) {
  sum((srate-(par[1]+par[2]*data[,1]+par[3]*(data[,1])^2+par[4]*(data[,1])^3))^2)
}

start3=c(0,0,0)
start4=c(0,0,0,0)
opt3 = optim(start3,sumSqMin3,data=grid,hessian=T,method="Nelder-Mead")
opt4 = optim(start4,sumSqMin4,data=grid,hessian=T,method="Nelder-Mead")

srquad = opt3$par[1]+opt3$par[2]*grid$grid+opt3$par[3]*(grid$grid)^2
srcubic = opt4$par[1]+opt4$par[2]*grid$grid+opt4$par[3]*(grid$grid)^2+opt4$par[4]*(grid$grid)^3

plot(grid$grid,srate,type = "l",
     main = "Plot of Spot Rates against Maturity",
     xlab = "Maturity of ZCB",
     ylab = "Spot Rates")
lines(grid$grid,srquad,col="red",lwd=2)
lines(grid$grid,srcubic,col="green",lwd=2)
legend("bottomright",c("Empirical","Quadratic","Cubic"),lty=c(1,1,1),col = c("black","red","green"))
```
(6)Comments
```{r}
options(scipen = 999)
opt3$value
opt4$value
```

Part 2
```{r}
#read the data
library(xts,quietly = T)
ZCBYF86=read.csv("C:/Users/Austin Lim/Desktop/SMU/AY2021/DSA302/Project/ZCBYF86.csv")
table(is.na(ZCBYF86))

#prepare the data
date=as.Date(ZCBYF86$Date,format="%d/%m/%Y")
ZCBYF86=cbind(date,ZCBYF86[,-1])
ZCBYF86.xts=xts(ZCBYF86[,-1],order.by=ZCBYF86[,1])
mat=as.numeric(c(1:30))
n=nrow(ZCBYF86.xts)
```
(1)Plot the data in a suitable way
```{r}
#rate against time: all bonds
library(ggplot2,quietly = T)
library(reshape2,quietly = T)
df <- melt(ZCBYF86 ,  id.vars = 'date', variable.name = 'series');
ggplot(df, aes(date,value)) + geom_line(aes(colour = series)) + labs(y="Yield (%)", x = "Time (Year)", title = "Yield Curves of ZCBYF86")

#rate against time: short term vs long term bonds
plot(ZCBYF86.xts[,c(1,30)],main="Short vs Long Term Bonds")

#rate against maturity
par(mfrow=c(1,2))
plot(mat,as.numeric(ZCBYF86.xts[1,]),type ="l",
     main = "1986-01-02",
     xlab = "Maturity", ylab = "Spot Rate")
plot(mat,as.numeric(ZCBYF86.xts["2000-08-01"]),type ="l",
     main = "2000-08-01",
     xlab = "Maturity", ylab = "Spot Rate")
par(mfrow=c(1,1))
```
(2)Fit the NS and NSS models to the yield data by minimizing sum of squared errors.
```{r}
# write a function to calculate the sum of squares
ns_sse <- function(par, maturity) {
  sum((yields-((par[1])^2 + (par[2] + par[3]/(par[4])^2)*(1-exp(-(par[4])^2*maturity))/((par[4])^2*maturity) - (par[3]*exp(-(par[4])^2*maturity)/(par[4])^2)))^2)
}
nss_sse <- function(par, maturity) {
  sum((yields-((par[1])^2 + (par[2] + par[3]/(par[4])^2)*(1-exp(-(par[4])^2*maturity))/((par[4])^2*maturity) - (par[3]*exp(-(par[4])^2*maturity)/(par[4])^2)+(par[5]/(par[6])^2*((1-exp(-(par[6])^2*maturity))/((par[6])^2*maturity)-exp(-(par[6])^2*maturity)))))^2)
}

# optimizing for 8650 days
theta_ns=data.frame()
theta_nss=data.frame()
error_ns=data.frame()
error_nss=data.frame()

for(i in 1:n){
  yields=as.numeric(ZCBYF86.xts[i,])
  ns_model = optim(c(1,1,1,1),ns_sse,maturity=mat,hessian=T,method="BFGS")
  nss_model = optim(c(1,1,1,1,1,1),nss_sse,maturity=mat,hessian=T,method="BFGS")
  theta_ns=rbind(theta_ns,ns_model$par)
  theta_nss=rbind(theta_nss,nss_model$par)
  error_ns=rbind(error_ns,ns_model$value)
  error_nss=rbind(error_nss,nss_model$value)
}

colnames(theta_ns)=c("beta0","theta1","theta2","beta3")
colnames(theta_nss)=c("beta0","theta1","theta2","beta3","theta4","beta5")
colnames(error_ns)="SSE_NS"
colnames(error_nss)="SSE_NSS"

theta_ns$theta0=(theta_ns$beta0)^2
theta_ns$theta3=(theta_ns$beta3)^2
theta_nss$theta0=(theta_nss$beta0)^2
theta_nss$theta3=(theta_nss$beta3)^2
theta_nss$theta5=(theta_nss$beta5)^2

#write to csv for convenience 
errors=cbind(error_ns,error_nss)
##write.csv(theta_ns,"theta_ns.csv",row.names = F)
##write.csv(theta_nss,"theta_nss.csv",row.names = F)
##write.csv(errors,"errors.csv",row.names = F)

#to use the csv files instead of optimizing again
##theta_nss=read.csv("C:/Users/Austin Lim/Desktop/theta_nss.csv")
##errors=read.csv("C:/Users/Austin Lim/Desktop/errors.csv")
##theta_ns=read.csv("C:/Users/Austin Lim/Desktop/theta_ns.csv")

# selection criteria 
aic_ns=n*log(errors$SSE_NS)+2*4;aic_nss=n*log(errors$SSE_NSS)+2*6
bic_ns=n*log(errors$SSE_NS)+log(n)*4;bic_nss=n*log(errors$SSE_NSS)+log(n)*6
criteria=as.data.frame(cbind(aic_ns,aic_nss,bic_ns,bic_nss));colnames(criteria)=c("AIC_NS","AIC_NSS","BIC_NS","BIC_NSS")

diff_sse=errors$SSE_NS-errors$SSE_NSS
diff_AIC=criteria$AIC_NS-criteria$AIC_NSS
diff_BIC=criteria$BIC_NS-criteria$BIC_NSS

par(mfrow=c(1,3))
plot(index(ZCBYF86.xts),diff_sse,type = "l",
     main = "Difference(SSE)",
     xlab = "Date", ylab = "Diff(SSE)");abline(h=0,col=2)
plot(index(ZCBYF86.xts),diff_AIC,type = "l",
     main = "Difference(AIC)",
     xlab = "Date", ylab = "Diff(AIC)");abline(h=0,col=2)
plot(index(ZCBYF86.xts),diff_BIC,type = "l",
     main = "Difference(BIC)",
     xlab = "Date", ylab = "Diff(BIC)");abline(h=0,col=2)
par(mfrow=c(1,1))
```
(3)What information could be extracted from theta?
```{r}
#filtering days where NS fitted better
indexes=which(errors$SSE_NSS>errors$SSE_NS)
filtered=ZCBYF86.xts[indexes,]
filtered_thetas=theta_ns[indexes,]

#theta0
plot(index(filtered),filtered_thetas$theta0,type="l",col=2,
     main = "Long Term Yield vs theta0 Comparison",
     xlab ="Date",ylab = "theta0 or long term spot rate")
lines(index(filtered),as.numeric(filtered[,30]))
legend("topright",c("theta0","SVEN30"),lty = c(1,1),col = c(2,1))

#theta1
steepness_par=filtered_thetas$theta1
steepness=as.numeric(sign(filtered[,2]-filtered[,1]))
plot(steepness_par,steepness,xlim=c(-50,50))
x=seq(-50,50,2)
y=NULL
for(i in 1:length(x)){
  if(x[i]<0){
    y[i]=1
  }
  else{
    y[i]=-1
  }
}
lines(x,y,col=2)

#verifying theta1
plot(mat,filtered[95,],type = "l",xlab = "maturity",ylab = "spot rate",ylim=c(8.5,9.5))
lines(mat,filtered[96,],col=2)
legend("topright",c("theta1<0","theta1>0"),lty = c(1,1),col = c(1,2))

#theta2/theta3
hump=filtered_thetas$theta2/filtered_thetas$theta3
hump_size=NULL
for(i in 1:nrow(filtered)){
  if(filtered[i,1]>filtered[i,2]){
    hump_size[i]=as.numeric(min(filtered[i,])-filtered[i,1])
  }
  else{
    hump_size[i]=as.numeric(max(filtered[i,])-filtered[i,1])
  }
}

hump_df=as.data.frame(cbind(hump[(hump>=-10)&(hump<=20)],hump_size[(hump>=-10)&(hump<=20)]));colnames(hump_df)=c("hump_par","hump_size")
ggplot(hump_df, aes(hump_par,hump_size)) +
  geom_point() +
  ylim(-1,5)+
  xlab("hump_par")+
  ylab("hump_size")+
  stat_smooth(method = lm, formula = y ~ poly(x, 2, raw = TRUE),se=F)

#verifying theta2/theta3
plot(mat,filtered[345,],type = "l",xlab = "maturity",ylab = "spot rate")
lines(mat,filtered[12,],col=2)
lines(mat,filtered[13,],col=3)
lines(mat,filtered[349,],col=4)
legend("bottomright",c("very negative","0-","0+","very positive"),lty = c(1,1,1,1),col = c(1,2,3,4))
```
Conclusions (Future Opportunities):
```{r}
#parameter stability
par(mfrow=c(2,2))
plot(index(ZCBYF86.xts),theta_ns$theta0, type="l", xlab="Date",ylab="theta0", main="theta0")
plot(index(ZCBYF86.xts),theta_ns$theta1, type="l", xlab="Date",ylab="theta1", main="theta1",ylim = c(-300,max(theta_ns$theta1)))
plot(index(ZCBYF86.xts),theta_ns$theta2, type="l", xlab="Date",ylab="theta2", main="theta2")
plot(index(ZCBYF86.xts),theta_ns$theta3, type="l", xlab="Date",ylab="theta3", main="theta3",ylim = c(0,400))
par(mfrow=c(1,1))

#finding volatilty of parameters?
library(dplyr)
theta_nss%>%
  summarise(sd0=sd(theta0),
            sd1=sd(theta1),
            sd2=sd(theta2),
            sd3=sd(theta3),
            sd4=sd(theta4),
            sd5=sd(theta5))
theta_ns%>%
  summarise(sd0=sd(theta0),
            sd1=sd(theta1),
            sd2=sd(theta2),
            sd3=sd(theta3))

library(xts)
library(MASS)
library(fGarch)
theta_nss.xts=xts(theta_nss[,c(7,2,3,8,5,9)],order.by = index(ZCBYF86.xts))

# for theta0 for example:
theta0=as.numeric(theta_nss.xts[,1])
m0_arch=garchFit(~garch(4,0),data=theta0,trace=F)
summary(m0_arch)
par(mfrow=c(2,2))
plot(index(ZCBYF86.xts),volatility(m0_arch),type="l",xlab="Year",ylab="Volatility",main="ARCH(4) Volatility, normal errors")
plot(index(ZCBYF86.xts),residuals(m0_arch,standardize=T),type="l",xlab="Year",ylab="Std Residual",main="ARCH(4) Std Res, normal errors")
qqnorm(residuals(m0_arch,standardize=T),datax=T,main="QQnorm of std residuals, normal errors")
qqline(residuals(m0_arch,standardize=T),datax=T,col="red")
pred=predict(m0_arch,n.ahead=100)
plot(1:100,pred$standardDeviation,type="l",xlab="Days",main="Predicted volatility over 100 days")
par(mfrow=c(1,1))
```